---
title: "NYPD Shooting Incident Data Report"
output:
  pdf_document: default
  html_document: default
---

## Summary
Which is more fatal, interracial incidents or non-interracial incidents?

Conclusion: Non-interracial incidents are more fatal.

## Importing Libraries
Importing packages for this project.

```{r import_packages, message=FALSE}
library(tidyverse)
library(lubridate)
```

This section is to import the data of NYPD Shooting Incident from its website.
```{r import_data, message=FALSE}
url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
datasets <- read_csv(url)
```

## Importing and Modifying Data
I tidy and transform the data.
```{r tidy_transform, message=TRUE}
tidy_datasets <- datasets %>%
  mutate(OCCUR_DATE = mdy(OCCUR_DATE), OCCUR_YEAR = year(OCCUR_DATE)) %>%
  select(INCIDENT_KEY, OCCUR_YEAR, OCCUR_DATE, OCCUR_TIME, BORO, STATISTICAL_MURDER_FLAG, PERP_AGE_GROUP, PERP_SEX, PERP_RACE, VIC_AGE_GROUP, VIC_SEX, VIC_RACE, LOCATION_DESC)
tidy_datasets
```

## Visualizing Data
### The Number of Shooting Incident
The number of shooting incidents in NY had been decreasing since 2006 until 2019. But 2020 saw a rebound probably because of the COVID-19 pandemic. 

```{r group_by_year}
year_datasets <- tidy_datasets %>%
  group_by(OCCUR_YEAR) %>%
  summarise(COUNT = n(), .groups = "drop") %>%
  select(OCCUR_YEAR, COUNT) %>%
  ungroup()

year_datasets %>%
  ggplot(aes(x = OCCUR_YEAR, y = COUNT)) +
  geom_line(aes(y = COUNT, color = "COUNT")) +
  geom_point(aes(y = COUNT, color = "COUNT")) +
  theme(legend.position = "right", axis.text.x = element_text(angle = 90, vjust = 1)) +
  scale_x_continuous(labels = year_datasets$OCCUR_YEAR, breaks = year_datasets$OCCUR_YEAR) +
  labs(title = "The Number of Shooting Incident in NY", x = "year", y = "the number of incident") +
  expand_limits(y = 0)
```
### The Number of Shooting Incidents by Race
Except "NA" and "UNKNOWN", the largest number of perpetrators and victims is "BLACK".

```{r group_by_race}
perp_race_datasets <- tidy_datasets %>%
  select(OCCUR_YEAR, PERP_RACE) %>%
  gather(OCCUR_YEAR, PERP_RACE) %>%
  group_by(OCCUR_YEAR, PERP_RACE) %>%
  count(PERP_RACE) %>%
  spread(PERP_RACE, n)

perp_race_datasets %>%
  gather(key, value, "AMERICAN INDIAN/ALASKAN NATIVE", "ASIAN / PACIFIC ISLANDER", "BLACK", "BLACK HISPANIC", "UNKNOWN", "WHITE", "WHITE HISPANIC", "<NA>") %>%
  ggplot(aes(x=OCCUR_YEAR, y=value, colour=key)) +
  geom_line() +
  theme(legend.position = "right", axis.text.x = element_text(angle = 90, vjust = 1)) +
  scale_x_continuous(labels = year_datasets$OCCUR_YEAR, breaks = year_datasets$OCCUR_YEAR) +
  labs(title = "The Number of Shooting Incident by Perpetrator's Race", x = "year", y = "the number of incident", colour = "race") +
  expand_limits(y = 0)
```
```{r group_by_vic_race}
vic_race_datasets <- tidy_datasets %>%
  select(OCCUR_YEAR, VIC_RACE) %>%
  gather(OCCUR_YEAR, VIC_RACE) %>%
  group_by(OCCUR_YEAR, VIC_RACE) %>%
  count(VIC_RACE) %>%
  spread(VIC_RACE, n)

vic_race_datasets %>%
  gather(key, value, "AMERICAN INDIAN/ALASKAN NATIVE", "ASIAN / PACIFIC ISLANDER", "BLACK", "BLACK HISPANIC", "UNKNOWN", "WHITE", "WHITE HISPANIC") %>%
  ggplot(aes(x=OCCUR_YEAR, y=value, colour=key)) +
  geom_line() +
  theme(legend.position = "right", axis.text.x = element_text(angle = 90, vjust = 1)) +
  scale_x_continuous(labels = year_datasets$OCCUR_YEAR, breaks = year_datasets$OCCUR_YEAR) +
  labs(title = "The Number of Shooting Incident by Victim's Race", x = "year", y = "the number of incident", colour = "race") +
  expand_limits(y = 0)
```
The number of incidents in-between race is larger than that of not. But the data does not have a sufficient property. It shows that almost 50% of data is "NA" between 2010 and 2020.
```{r same_race_incident}
same_race_datasets <- tidy_datasets %>%
  mutate(IN_BETWEEN_RACE = PERP_RACE == VIC_RACE) %>%
  select(OCCUR_YEAR, IN_BETWEEN_RACE) %>%
  count(OCCUR_YEAR, IN_BETWEEN_RACE) %>%
  group_by(OCCUR_YEAR) %>%
  mutate(pct= prop.table(n) * 100)

same_race_datasets %>%
  ggplot(aes(fill=IN_BETWEEN_RACE, y=pct, x=OCCUR_YEAR)) + 
  geom_bar(position="fill", stat="identity")
```

### Modeling (interracial and non-interracial fatality rate)
I created a model of interracial and non-interracial fatality rate.
```{r modeling}
per_thou_datasets <- tidy_datasets %>%
  filter(!is.na(VIC_RACE)) %>%
  filter(!is.na(PERP_RACE)) %>%
  filter(VIC_RACE == PERP_RACE) %>%
  group_by(OCCUR_YEAR) %>%
  summarise(COUNT = n(), FATAL_COUNT = sum(STATISTICAL_MURDER_FLAG == TRUE), .groups = "drop") %>%
  mutate(IN_BETWEEN_PER_THOU = FATAL_COUNT / COUNT * 1000)

per_thou_interrace_datasets <- tidy_datasets %>%
  filter(!is.na(VIC_RACE)) %>%
  filter(!is.na(PERP_RACE)) %>%
  filter(VIC_RACE != PERP_RACE) %>%
  group_by(OCCUR_YEAR) %>%
  summarise(COUNT = n(), FATAL_COUNT = sum(STATISTICAL_MURDER_FLAG == TRUE)) %>%
  mutate(INTERRACE_PER_THOU = FATAL_COUNT / COUNT * 1000)

joined <- per_thou_interrace_datasets %>%
  full_join(per_thou_datasets, by = "OCCUR_YEAR") %>%
  mutate(RATIO = IN_BETWEEN_PER_THOU / INTERRACE_PER_THOU)

mod <- lm(FATAL_COUNT ~ COUNT, data = per_thou_datasets)
summary(mod)

interrace_mod <- lm(FATAL_COUNT ~ COUNT, data = per_thou_interrace_datasets)
```

```{r prediction}
pred <- per_thou_datasets %>%
  mutate(pred = predict(mod))

interrace_pred <- per_thou_interrace_datasets %>%
  mutate(pred = predict(interrace_mod))

pred %>%
  ggplot() +
  geom_point(aes(x = COUNT, y = FATAL_COUNT), color = "blue") +
  geom_point(aes(x = COUNT, y = pred), color = "red")

interrace_pred %>%
  ggplot() +
  geom_point(aes(x = COUNT, y = FATAL_COUNT), color = "blue") +
  geom_point(aes(x = COUNT, y = pred), color = "red")
```

Lastly, I calculate the ratio, non-interracial divided by interracial. if ratio is more than 1.0, it means that non-interracial is more fatal. If ratio is less than 1.0, interracial is more fatal.

The red dots are actual.
The blue line is a prediction.
The black line is horizontally 1.0.

As the graph shows, the red dots are almost alwaus above the black line. In addition, the prediction line is also above 1.0. In conclusion, non-interracial incidents are more fatal.

```{r datasets}

joined %>%
  ggplot(aes(OCCUR_YEAR, RATIO)) +
  geom_point(color = "red") +
  stat_smooth(method = lm) +
  geom_hline(yintercept = 1)
```